HEP Software Foundation Community White Paper Working Group -- Data Processing Frameworks

### 2018-10-11 CWP/papers/HSF-CWP-2017-08_framework/latex/df_fwk_summary.tex
Add frameworks paper abstract
Correct some small mistakes in the text as well

\abstract{Abstract goes here.}

@@

\title{HEP Software Foundation Community White Paper -- Data Processing Framework WG Report}

\author{HEP Software Foundation:}
\author[d]{Paolo Calafiura}
\author[a,1]{Benadikt Hegner}
\author[c]{Chris Jones}
\author[c,1]{Jim Kowalkowski}
\author[c,1]{Elizabeth Sexton-Kennedy}
\author[b]{Michel Jouvin}
\author[d]{Several Others}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{Fermi National Accelerator Laboratory, Batavia, Illinois, USA}
\affiliation[d]{Lawence Berkeley National Laboratory, Berkeley, CA, USA}
\affiliation[e]{Other Places}
\affiliation[1]{Paper Editor}

@@

The time periods of interest for this document are DUNE and HL-LHC,
which will deliver on the order of 50 PB data events per year per
experiment. The results of the proposed R\&D ought to be used for
building the final software systems that will be utilized in
commissioning and operations of these experiments and the processing
of the data.

@@

\begin{enumerate}
\item
    Changes needed in the programming model that are necessary to
    handle the massive parallelism that will be present throughout all
    layers in the available compute facilities. This is necessary
    because of the ever-increasing availability of specialized compute
    resources, and includes GPGPUs, Tensor Processing Units (TPUs),
    and tiered memory systems integrated with storage, and ultra
    high-speed network interconnects.
\item
    Challenges related to advanced detector technology, like finer
    granularity, high bandwidth continuous readout DAQ, and the need
    for a very tight feedback loop for conditions, which in turn
    requires a focus on low-latency solutions.
\item
    Tighter integration with the computing model: informing
    higher-level systems (workflow, data management, workload
    management, job management) based on fundamental changes in
    compute facilities.
\item
    Support structures that permit frameworks to be collaboratively
    developed and maintained across a large number of
    experiments. This includes excellent integration with the wider
    ecosystem encompasses development, deployment, and runtime
    components.
\item
    Provide flexibility and the interfaces required to efficiently
    integrate the various parallelization efforts on simulation,
    reconstruction and other fields.
\end{enumerate}

@@

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. A complete description
of framework use cases written for the Gaudi collaboration is
described in {[}2.fixme{]} and these are sufficiently general to apply
to all HEP experiments. The Gaudi framework {[}70{]} was originally
developed by LHCb, but is also used by ATLAS and various non-LHC
experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
{[}71{]} which was forked to provide the art framework for the Fermilab
Intensity Frontier experiments~\cite{Green:2012gv},{[}72{]}.  Belle II uses
basf2~\cite{1742-6596-331-3-032024},{[}73{]}. The linear collider community
developed and uses Marlin~\cite{Gaede:2006pj},{[}74{]}. The FAIR experiments
use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
and ALICE are now developing a new framework, which is called
O2~\cite{O2},{[}75{]}. At the time of writing, most major frameworks support
basic parallelisation, both within and across events, based on a
task-based model~\cite{Jones:2015soc}~\cite{Clemencic:2015paa}. O2 already 
includes additional multi-node setups and communication.

The frameworks provide the necessary functionality like I/O,
scheduling, configuration, logging, etc to support the execution of
these processing components. The mentioned components provide
functionalities like pattern finding in a certain sub-detector or the
high-level identification of a given particle type. This layout allows
independent development and a high flexibility in the usage of physics
algorithms within an experiment collaboration.

@@

Current practice for throughput-maximizing system architectures have
constrained the scope of framework designs. Framework applications
have largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this
tradition starting in the beginning of Run 2, by utilizing all cores
on one virtual node in one process space using threading. ATLAS is
currently using a multi-process fork-and-copy-on-write solution to
remove the constraint of one core/process, and is now moving to the 
multithreading approach too. Both experiments were
driven to solve this problem, by the ever growing needs for more
memory per process brought on by the increasing complexity of LHC
events. Current practice manages system-wide (or facility-wide)
scaling by dividing up datasets, generating a framework application
configuration, and scheduling jobs on nodes/cores to consume all
available resources. Given anticipated changes in hardware
(heterogeneity, connectivity, memory, storage) available at large
computing facilities, the interplay between workflow/workload
management systems and framework applications need to be carefully
examined. It may be advantageous to permit framework applications (or
systems) to span resources, permitting them to be first-class
participatents in the business of scaling within a facility. O2 provides
a successful proof-of-principle of this approach.

\section{Roadmap}
\label{sec:roadmap}
Forward-looking work is underway as part of projects funded through government agencies, laboratories, and collaborations.  We want to be sure that relevant ideas and accomplishment are known, and that the groups doing this work have a place to report to and receive feedback for everyone’s benefit. 
To organize the community, one needs to establish regular working group meetings, on a bi-monthly basis as we did with the concurrency forum. Face to face workshops after at least the 1st and the 3rd year can be co-hosted with events like CHEP and/or the WLCG workshops. A future planning workshop for transforming the results of the R&D activities into a full development and deployment project plan should happen at the 5-year timescale.

@@

\paragraph{Domain-specific language} We believe it is useful to
describe the language that HEP uses to define and describe how to
simulate and reconstruct physics events. By language, we mean the
terms used to communicate and express how tasks are described and
carried out within a framework. This includes not only expressing data
dependencies, but also resource preferences and constraints, such as
GPU. The goal here is to provide enough information for a group to
take on development of domain-specific libraries components and tools
that will increase the efficiency of carrying out physics. A good example
is how ML toolkits have evolved over the past few years. The 
abstractions that have been developed have greatly increased productivity 
and growth in the ML space such as the abstractions in Tensor Flow that 
allow a coding of the matrix algebra that then gets remapped internally to 
match the shape of the data being operated on.  The user only has take care
of getting the domain science functions correct.

>>>>

\abstract{Data processing frameworks are an essential part of HEP
experiments' software stacks. Frameworks provide a means by which code
developers can undertake the essential tasks of physics data processing,
accessing relevant inputs and storing their outputs, in a coherent way without
needing to know the details of other domains. Frameworks provide essential
core services for developers and help deliver a configurable working
application to the experiments' production systems.
Modern HEP processing frameworks are in the process of adapting to a new
computing landscape dominated by parallel processing and heterogeneity,
which pose many questions regarding enhanced functionality and scaling
that must be faced without compromising the maintainability of the code.
In this paper we identify a program of work that can help further clarify
the key concepts of frameworks for HEP and then spawn R\&D activities that
can focus the community's efforts in the most efficient manner to address
the challenges of the upcoming experimental program.
}

@@

\title{HEP Software Foundation Community White Paper Working Group -- Data Processing Frameworks}

\author{HEP Software Foundation:}
\author[d]{Paolo Calafiura}
\author[a,1]{Benadikt Hegner}
\author[c]{Chris Jones}
\author[b]{Michel Jouvin}
\author[c,1]{Jim Kowalkowski}
\author[c,1]{Elizabeth Sexton-Kennedy}
\author[a]{Graeme A Stewart}
\author[d]{Several Others - Charles, Marco, ?}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{Fermi National Accelerator Laboratory, Batavia, Illinois, USA}
\affiliation[d]{Lawrence Berkeley National Laboratory, Berkeley, CA, USA}
\affiliation[e]{Other Places}
\affiliation[1]{Paper Editor}

@@

The time periods of interest for this document are DUNE and HL-LHC,
which will deliver on the order of 50 PB of data events per year per
experiment. The results of the proposed R\&D ought to be used for
building the final software systems that will be utilized in
commissioning and operations of these experiments and the processing
of the data.

@@

\begin{enumerate}
\item
    Changes needed in the programming model that are necessary to
    handle the massive parallelism that will be present throughout all
    layers in the available computing facilities. This is necessary
    because of the ever-increasing availability of specialized compute
    resources, including GPGPUs, Tensor Processing Units (TPUs),
    tiered memory systems integrated with storage, and ultra
    high-speed network interconnects.
\item
    Challenges related to advanced detector technology, like finer
    granularity, high bandwidth continuous readout DAQ, and the need
    for a very tight feedback loop for conditions, which in turn
    requires a focus on low-latency solutions.
\item
    Tighter integration with the computing model: informing
    higher-level systems (workflow, data management, workload
    management, job management) based on fundamental changes in
    compute facilities.
\item
    Support structures that permit frameworks to be collaboratively
    developed and maintained across a large number of
    experiments. This includes excellent integration with the wider
    ecosystem encompasses development, deployment, and runtime
    components.
\item
    Provide flexibility and the interfaces required to efficiently
    integrate the various parallelization efforts on simulation,
    reconstruction and other fields.
\end{enumerate}

@@

Although most frameworks used in HEP share common concepts, there are,
for mainly historical reasons, a number of different implementations;
some of these are shared between experiments. A complete description
of framework use cases written for the Gaudi collaboration is
described in {[}2.fixme{]} and these are sufficiently general to apply
to all HEP experiments. The Gaudi framework {[}70{]} was originally
developed by LHCb, but is also used by ATLAS and various non-LHC
experiments. CMS uses its own CMSSW framework~\cite{Bayatian:2006nff},
{[}71{]} which was forked to provide the art framework for the Fermilab
Intensity Frontier experiments~\cite{Green:2012gv},{[}72{]}.  Belle II uses
basf2~\cite{1742-6596-331-3-032024},{[}73{]}. The linear collider community
developed and uses Marlin~\cite{Gaede:2006pj},{[}74{]}. The FAIR experiments
use FairROOT, closely related to ALICE's AliROOT. The FAIR experiments
and ALICE are now developing a new framework, which is called
O2~\cite{O2},{[}75{]}. At the time of writing, most major frameworks support
basic parallelisation, both within and across events, based on a
task-based model~\cite{Jones:2015soc,Clemencic:2015paa}. O2 already
includes additional multi-node setups and communication.

The frameworks provide the necessary functionality like I/O,
scheduling, configuration, logging, etc.\ to support the execution of
these processing components. The aforementioned components provide
functionalities like pattern finding in a certain sub-detector or the
high-level identification of a given particle type. This layout allows
independent development and a high flexibility in the usage of physics
algorithms within an experiment collaboration.

@@

Current practice for throughput-maximizing system architectures have
constrained the scope of framework designs. Framework applications
have largely been viewed by the system as a batch job with complex
configuration, consuming resources according to rules dictated by the
computing model: one process using one core on one node, operating
independently with a fixed size memory space on a fixed set of files
(streamed or read directly). Only recently has CMS broken this
tradition starting in the beginning of Run 2, by utilizing all cores
on one virtual node in one process space using threading. ATLAS is
currently using a multi-process fork-and-copy-on-write solution to
remove the constraint of one core/process, and is now moving to the
multithreading approach too. Both experiments were
driven to solve this problem by the ever growing needs for more
memory per process brought on by the increasing complexity of LHC
events. Current practice manages system-wide (or facility-wide)
scaling by dividing up datasets, generating a framework application
configuration, and scheduling jobs on nodes/cores to utilize all
available resources. Given anticipated changes in hardware
(heterogeneity, connectivity, memory, storage) available at large
computing facilities, the interplay between workflow/workload
management systems and framework applications needs to be carefully
examined. It may be advantageous to permit framework applications (or
systems) to span resources, permitting them to be first-class
participatents in the business of scaling within a facility. O2 provides
a successful proof-of-principle of this approach.

\section{Roadmap}
\label{sec:roadmap}
Forward-looking work is underway as part of projects funded through government agencies, laboratories, and collaborations.  We want to be sure that relevant ideas and accomplishment are known, and that the groups doing this work have a place to report to and receive feedback for everyone’s benefit.
To organize the community, one needs to establish regular working group meetings, on a bi-monthly basis as we did with the concurrency forum. Face to face workshops after at least the 1st and the 3rd year can be co-hosted with events like CHEP and/or the WLCG workshops. A future planning workshop for transforming the results of the R\&D activities into a full development and deployment project plan should happen at the 5-year timescale.

@@

\paragraph{Domain-specific language} We believe it is useful to
describe the language that HEP uses to define and describe how to
simulate and reconstruct physics events. By language, we mean the
terms used to communicate and express how tasks are described and
carried out within a framework. This includes not only expressing data
dependencies, but also resource preferences and constraints, such as
GPUs. The goal here is to provide enough information for a group to
take on development of domain-specific libraries components and tools
that will increase the efficiency of carrying out physics. A good example
is how ML toolkits have evolved over the past few years. The
abstractions that have been developed have greatly increased productivity
and growth in the ML space such as the abstractions in Tensor Flow that
allow a coding of the matrix algebra that then gets remapped internally to
match the shape of the data being operated on.  The user only has take care
of getting the domain science functions correct.
